---
title: "bike-analysis"
author: "Steven"
date: "2023-03-08"
output:
  html_document: default
---

# Cyclistic bike Analysis

#### Step 1: Setting up environment with required packages.

```{r}
#Set work directory to where the file is located
setwd("E:/data analytic/case_study_bike/bike_data/2020-2021")

#Installing and loading tidyverse package
install.packages("tidyverse")
library(tidyverse)
```

#### Step 2: Storing the merged CSV into a dataframe

```{r}
df <- read_csv("merged_data/merged.csv")
str(df)
```

#### Step 3: Clean up and Add data to Prepare for Analysis

```{r}
# Inspect data
colnames(df)
nrow(df)
head(df)
```

##### Checking "member_casual" column.

```{r}
table(df$member_casual)
```

##### Add columns that list the date, month, day, and year of each ride.

##### *Add Columns for start date.*

```{r}
df$start_date <- as.Date(df$started_at)
df$start_month <- format(df$start_date, "%m")
df$start_day <- format(df$start_date, "%d")
df$start_year <- format(df$start_date, "%Y")
df$start_weekname <- format(df$start_date, "%A")
df$start_time <- format(df$started_at, "%H:%M")
```

##### *Add Columns for end date.*

```{r}
df$end_date <- as.Date(df$ended_at)
df$end_month <- format(df$end_date, "%m")
df$end_day <- format(df$end_date, "%d")
df$end_year <- format(df$end_date, "%Y")
df$end_weekname <- format(df$end_date, "%A")
df$end_time <- format(df$ended_at, "%H:%M")
```

##### This allows us to aggregate ride data in a deeper level.

##### Adding "duration" column, with difference between starting and ending time calculated.

```{r}
df$duration <- difftime(df$ended_at, df$started_at)

# Inspect new column
summary(df$duration)


```

##### Change 'duration' column to numeric for further calculations on the data.

```{r}
df$duration <- as.numeric(as.character(df$duration))
summary(df$duration)
```

##### Several hundred bikes were took out of circulation for Quality Control reasons and have caused some negative values in "duration" column.

##### Filter out "bad" data and store into df2.

```{r}
df2 <- df[!(df$start_station_name=="HQ QR" | df$duration < 0),]

# Check "duration" column in df2
summary(df2$duration)
```

#### Inspect new dataframe

```{r}
head(df2)
```

### Step 4: Conduct Descriptive Analysis

##### Summary statistics on ride duration (in seconds).

```{r}
summary(df2$duration)
```

##### Compare members and casual users

```{r}
mean_dur_table <- aggregate(df2$duration ~ df2$member_casual, FUN = mean)
median_dur_table <- aggregate(df2$duration ~ df2$member_casual, FUN = median)
max_dur_table <- aggregate(df2$duration ~ df2$member_casual, FUN = max)
min_dur_table <- aggregate(df2$duration ~ df2$member_casual, FUN = min)

# change label names
colnames(mean_dur_table) <- c("Member Type", "Mean duration")
colnames(median_dur_table) <- c("Member Type", "Median duration")
colnames(max_dur_table) <- c("Member Type", "Max duration")
colnames(min_dur_table) <- c("Member Type", "Min duration")

# Print output
print(mean_dur_table)
print(median_dur_table)
print(max_dur_table)
print(min_dur_table)
```

##### Average riding time by each day of week for *Casual Users* and *Members*.

```{r}

df2$start_weekname <- ordered(df2$start_weekname, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(df2$duration~df2$member_casual + df2$start_weekname, FUN = mean)
```

##### Analyze ridership data by type and weekday

```{r}
export1 <- df2 %>%  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(duration))

export1
```

##### Visualize the number of rides by type

```{r}
 weekdays <- df2 %>%  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(duration)) %>% 
  arrange(member_casual, weekday) %>% na.omit()

 weekdays %>% ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge")
```

##### Visualization for average duration

```{r}
weekdays %>% ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) + geom_col(position = "dodge")
```


##### Visualization for peak hours
```{r}


peak_hours <- df2 %>% 
  mutate(start_hour = format(started_at,"%H")) %>% 
  group_by(member_casual,start_hour) %>% 
  summarise(number_of_rides = n(),average_duration = mean(duration)) %>% 
  na.omit()

peak_hours %>% ggplot(aes(x = start_hour, y = average_duration, fill = member_casual)) + geom_col(position = "dodge")+ labs(title = "Average duration of ride by each Hour", x="Hour of Day", y = "Average Duration (seconds)", fill = "Member type") 
  
```
```{r}
peak_hours %>% ggplot(aes(x = start_hour, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge")+ labs(title = "Number of rides by each Hour", x="Hour of Day", y = "Number of Rides", fill = "Member type") 
```

##### Look at which start / end station is most used

```{r}
export2 <- df2 %>%
  group_by(member_casual, start_weekname) %>% na.omit() %>% 
  summarise(max_start_station = names(which.max(table(start_station_name))), max_end_station = names(which.max(table(end_station_name))), start_lat = mean(start_lat), start_lng = mean(start_lng), end_lat = mean(end_lat), end_lng = mean(end_lng), .groups = "keep")

export3 <- df2 %>%
  group_by(member_casual, format(started_at,"%H")) %>% na.omit() %>% 
  summarise(max_start_station = names(which.max(table(start_station_name))), max_end_station = names(which.max(table(end_station_name))), start_lat = mean(start_lat), start_lng = mean(start_lng), end_lat = mean(end_lat), end_lng = mean(end_lng), .groups = "keep")

export3
```



```{r}
df2 %>% aggregate(end_station_name ~ member_casual + format(started_at,"%H"), FUN=max)
```



### Step 5: Export Summary File for Further Analysis

```{r}
write.csv(export1, file = "export_data/summary1.csv")
write.csv(export2, file = "export_data/most_used_station_byweek.csv")
write.csv(export3, file = "export_data/most_used_station_byhour.csv")
write.csv(peak_hours, file = "export_data/peak_hour.csv")

```

